# Importing and Viewing Data

## Example: House Price Data

The data that we will be using in this session come from R's `wooldridge` package, which includes 115 data sets from "Introductory Econometrics: A Modern Approach, 7e" by Jeffrey M. Wooldridge".

We will start by using `hprice2` data, which includes information on 506 communities in the US Boston area. Below table provides a list of variables that we will use:

| Variable | Description                                               |
|----------|-----------------------------------------------------------|
| price    | Median housing price, \$                                  |
| crime    | Crimes committed per capita                               |
| nox      | Nitrogen Oxide in the air, in parts per million           |
| stratio  | Average student-teacher ratio of schools in the community |
| rooms    | Average number of rooms in houses in the community        |
| lprice   | Logarithm of price                                        |

## Packages and libraries

In order to access specialised data analysis tools in R, we will need to install some R packages.

"An R **package** is a collection of functions, data, and documentation that extends the capabilities of base R.

We will start by installing the `wooldridge` package. We require this package to access the Wooldridge data sets, mentioned above.

```{r}
# install.packages("wooldridge")
```

To install `wooldridge` package, type the above line of code in the console (without the `#`), and then press enter to run it. R will download the packages from CRAN and install them on to your computer.

Once installed, you may use this package after loading it with the `library()` function.

```{r}
library(wooldridge)
```

## Open data as an R object

The `hprice2` data that we will be using already comes in R format within the `wooldridge` package. We can call it to our environment by using the `data()` function.

```{r}
data("hprice2")
```

Observe that `hprice2` is now added as an object to your environment. We can use the `View` function to view it (Note the capital V in `View`).Type the below line without the `#` in your console.

```{r}
# View(hprice2)
```

## Import Excel data into R

Most data we work with are initially in Excel or in text (such as csv) format. Importing data is one of the rare R commands that you can proceed with using the menu. You may find the `Import Dataset` button on the top right window and also under File menu. The `hprice2` data is provided to you in Excel format. Try to import it using `Import Dataset` but this time, in the opening window, name the dataset as `df` under `Import Options` towards the left bottom.

You will see that R executes the following three lines of code:

```{r}
library(readxl)
df <- read_excel("data/hprice2.xlsx")
View(df)
```

The first line call the library required; the second line reads the Excel file and saves it under name `df` and the third line views the newly imported data.

You may use `head()` and `tail()` functions to view, respectively, the first few and the last few lines of the data.

```{r}
# View the first few observations of the data
head(hprice2)
```

```{r}

# View the last few observations of the data
tail(hprice2)
```
