# Basic Plots of Data

```{r, echo=FALSE}
library(wooldridge)
data("hprice2")
```

In Session 2 of this workshop, we will be using `ggplot2` for data visualisation in R, which produces better-looking plots. For the moment, below are a few examples of what we can do with base R (with base R, we refer to the operations of R that could be done without installing or calling additional libraries).

**Scatter plot**

We may use scatter plots to depict the relationship between two variables. In the example below, we are depicting the relationship between housing price and crime. You may read the `~` between the two variables as *approximated by*. So here, price is approximated by crime.

```{r}
# scatterplot showing the relationship between price and crimes 
plot(hprice2$price ~ hprice2$crime)

```

We may add more informative titles to our axes and also provide a general title for the plot:

```{r}
# scatterplot showing the relationship between price and crimes 
plot(hprice2$price ~ hprice2$crime, 
     xlab = "Crimes committed per capita", 
     ylab = "Median housing price, $", 
     main = "Negative effect of crime on housing price")
```

As expected, there is a negative relationship between the crime rates in a location and the median house prices in that location.

In the following pages, we will be estimating regresions to model the houseing price with a set of independent variables. Before moving on to regression analysis, it is good practice to get to know our dependent variable. A histogram will reveal its distribution.

**Histogram**

Below, we provide histograms of `price` and `lprice` (logarithmic price). Which one should we use in our regression?

```{r}
# Histogram of price and logarithmic price
hist(hprice2$price, 
     xlab = "Median housing price, $", 
     main = "Histogram of median housing price, $")
```

```{r}
hist(hprice2$lprice, 
     xlab = "Logarithm of median housing price, $", 
     main = "Logarithm of median housing price, $")

```
