# Basic Plots of Data

In Session 2 of this workshop, we will be using `ggplot2` for data visualisation in R, but for the moment, below are a few examples of what we can do with base R (with base R, we refer to the operations of R that could be done without installing or calling additional libraries).

**Scatterplot** of

```{r}
# scatterplot showing the relationship between price and crimes 
plot(hprice2$price ~ hprice2$crime, xlab = "Crimes committed per capita", 
     ylab = "Median housing price, $", 
     main = "Negative effect of crime on housing price")

# Histogram of price and logarithmic price
hist(hprice2$price, xlab = "Median housing price, $", 
     main = "Histogram of median houseing price, $")

hist(hprice2$lprice, xlab = "Logarithm of median housing price, $", 
     main = "Logarithm of median housing price, $")

```

here are alternative packages which could be used to label variables. We will be using the `Hmisc` package. Remember to install this package and call the library by running the code below (without the `#`):

```{r}
# install.packages("Hmisc") # to label variables
library(Hmisc)
```

We are ready to label our variables:

```{r}
# label variables
label(hprice2$price) <- "Median housing price, $"
label(hprice2$crime) <- "Crimes committed per capita"
label(hprice2$nox) <- "Nitrogen Oxide in the air, in parts per million"
label(hprice2$stratio) <- "Average student-teacher ratio of schools in the community"
label(hprice2$rooms) <- "Average number of rooms in houses in the community"
```
